Preschoolers form conventional pacts with each other to communicate about novel shapes

### 2025-01-14 write-ups/kid_tangrams_cogsci_2025/kid_tangrams_cogsci_2025.Rmd
start of V edit pass

---
title: "TODO Preschoolers form conventional pacts with each other to communicate about novel shapes"
bibliography: preschool-tangrams.bib
csl: apa7.csl
document-params: "10pt, letterpaper"

author-information: > 
    \author{{\large \bf Veronica Boyce (vboyce@stanford.edu)} \\ Department of Psychology, \\Stanford University \And {\large \bf Robert Z. Sparks (bsparks@stanford.edu) } Department of Psychology, \\Stanford University
    \AND {\large \bf Yannick Mofor (TODO email)} \\ TODO affiliation, Stanford University \And {\large \bf Michael C. Frank (mcfrank@stanford.edu)} \\ Department of Psychology, \\ Stanford University}

abstract: >
   Learning language requires learning not only the content of language, but also skills that allow us to use language to communicate. Iterated reference games provide a window into such skills, requiring rich communication as participants jointly converge on mutually understandable names for initially novel objects. Some classical experiments with young children are interpreted as showing that 4-5-year old children are incapable of succeeding at iterated reference games. Here, we revisit the question of young children's referential communicative abilities using a simpler, child-friendly paradigm. Across 51 pairs of 4-5-year old children, we find that preschool-aged children were successful in an iterated reference game. Children were Y% accurate, and they often used descriptions similar to their partner's. These findings suggest that children’s capacity to construct effective referring expressions in novel contexts emerges earlier than once thought, consistent with a larger picture of the early emerging ability to pragmatically use language in communication. TODO fix wrap up sentence! TODO possibly shorten!

keywords: >
    development; iterated reference game; TODO keywords

output: cogsci2024::cogsci_paper
final-submission: \cogscifinalcopy
---

@@

(Note: we should talk about whether it makes more sense to do all the methods and then all the results versus the current expt 1 than expt 2 layout.)

Our goal for experiment 1 was to test young children's ability to coordinate on descriptions to abstract shapes that their partner could understand. Young children can be very sensitive to task demands and cognitive load that can hide early abilities [CITATION WANTED], so we adapted the experimental framework from @leung2024, but further simplified it by reducing the total pool of targets and the number of trials children completed.

This experiment was pre-registered at https://osf.io/kcv8j.

@@

Statistical analyses were run in TODO CITE brms with weakly informative priors. We present the estimate and 95% credible intervals.

@@

[less important, could drop] As another measure of children's performance, we looked at how long each trial took to see if children were getting faster over time. We ran a Bayesian mixed effects model of how long each trial took over time: `r speed_form |> form()`. The first critical trial averaged `r stats_text(speed_mod_1, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_1, 2)`).

@@

In iterated reference games in adults, a canonical finding is that the length of descriptions goes from long to short over repeated references to the initially hard to describe shapes [@clark1986; @hawkins2020b; @boyce2024]. We looked at how long the descriptions the children used were to see if the same trend occurred. 
We ran a Bayesian mixed effects model of how long of a description the "teller" produced: `r description_form |> form()`. The initial length was `r stats_text(description_mod_1, 1)` and description length was relatively stable over time (`r stats_text(description_mod_1, 2)`, shown in Figure \ref{length}). How long of a description is initially warranted depends on how iconic or easily describable the shape is, as well as how large and close the contrast set is. In this case, the low number of options to distinguish and relatively easy to describe shapes may mean long initial descriptions are less necessary. (The adult controls in @leung2024 use shorter initial descriptions than adults in studies with larger arrays of harder to distinguish images.) However, young children may also choose to provide shorter descriptions than adults would, and young children may accept shorter descriptions as adequate whereas adults may ask for more information to increase their confidence level before making a guess.

@@

While description length is used as a proxy for measuring convention formation, better measurements for convergence look at the actual content of the utterances, which does not always track description length [@boyce2024]. Following @boyce2024, we use Sentence-BERT to embed the descriptions in a semantic vector space, and then use the cosine between embeddings as a measure of semantic similarity between descriptions. We compared the semantic similarities between descriptions of the same target based on who produced the description. Our question was whether the two children within the same game produced descriptions that are more similar to each other than two children in different games.

@@

In the first experiment, we adapted the paradigm of @leung2024 for pairs of children, taking an already simple set up and making it shorter. Our goal was to see if young children were at all able to provide descriptions to each other adequate to identify abstract figures. so children received a lot of scaffolding around the experimental interaction. Sometimes, this included experimenters echoing children's descriptions which could potentially influence children's responses. In Experiment 2, we repeated the same paradigm, with tighter experimenter controls and a larger sample size.

@@

```{=latex}

\begin{table}
	\caption{Example descriptions children used for different tangrams. TODO may want to shorter or pair with the tangram images}
		\label{example}
	\begin{tabular}[H]{p{22em}}
		Descriptions \\
		\hline
		%hold
		"Hold":\\
		$\bullet$ person\\
		$\bullet$ a person holding a sandwich\\
		$\bullet$ a people carrying a box of dirt\\
		$\bullet$ a monster\\
		$\bullet$ someone holding a plate and giving it to a restaurant and has watermelon\\
		\hline
	"Jump":\\
		$\bullet$ vampire\\
		$\bullet$ hopping\\
		$\bullet$ a person flying\\
		$\bullet$ somebody skydiving, not in the airplane\\
		$\bullet$ a person\\
		$\bullet$ a kite\\
		$\bullet$ a triangle with a head on it with feet\\
		\hline
		"Swim":\\
		$\bullet$ racecar\\
		$\bullet$ airplane\\
		$\bullet$ alligator\\
		$\bullet$ a person fell down\\
		$\bullet$ a boat\\
		$\bullet$ a person that's in a race car that has one triangle and two triangles\\
		\hline
		"Walk":\\
		$\bullet$ person\\
		$\bullet$ a person walking\\
		$\bullet$ a person looking down\\
		$\bullet$ a people, but it doesn't have any arms\\
		\hline
	\end{tabular}
\end{table}

@@

# Post hoc results

## Meta-analytic placeholder

Might want to split out and incorporate in along with expt 2 analyses?

@@

@@

Query -- I did not include experiment # as a random effect here, and maybe should have? I can rerun if we decide we want this analysis and think expt # as a random effect with game nested within it would be useful.

## Accuracy is predictive of similarity to next

@@

Across 2 experiments and 51 pairs of 4 and 5-year old children, we tested how well children were at producing referential expressions to allow their partner to find a matching abstract shape. Children varied substantially in what sorts of descriptions they produced, but overall accuracy was high (NUMBER), indicating that children were generally able to produce adequate descriptions. Additionally, children's utterances showed signs of converging toward conceptual pacts.  While this task is substantially scaled down relative to measures used for adult competence, and children do not display all the typical trends (i.e no sign of shortening of referential expression), it does suggest that the relevant communication skills are present at least in rudimentary form by the end of the preschool years.

>>>>

---
title: "Preschoolers form conventional pacts with each other to communicate about novel shapes"
bibliography: preschool-tangrams.bib
csl: apa7.csl
document-params: "10pt, letterpaper"

author-information: > 
    \author{{\large \bf Veronica Boyce (vboyce@stanford.edu)} \\ Department of Psychology, \\Stanford University \And {\large \bf Robert Z. Sparks (bsparks@stanford.edu) }\\ Department of Psychology, \\Stanford University
    \AND {\large \bf Yannick Mofor (yannickm@stanford.edu)} \\ Department of Computer Science, \\ Stanford University \And {\large \bf Michael C. Frank (mcfrank@stanford.edu)} \\ Department of Psychology, \\ Stanford University}

abstract: >
   Learning language requires learning not only the content of language, but also skills that allow us to use language to communicate. Iterated reference games provide a window into such skills, requiring rich communication as participants jointly converge on mutually understandable names for initially novel objects. Some classical experiments with young children are interpreted as showing that 4-5-year old children are incapable of succeeding at iterated reference games. Here, we revisit the question of young children's referential communicative abilities using a simpler, child-friendly paradigm. Across 51 pairs of 4-5-year old children, we find that preschool-aged children were successful in an iterated reference game. Children were 85% accurate, and they often used descriptions similar to their partner's. These findings suggest that children’s capacity to construct effective referring expressions in novel contexts emerges earlier than once thought, consistent with a larger picture of the early emerging ability to pragmatically use language in communication.

keywords: >
    development; iterated reference game; pragmatics; preschoolers

output: cogsci2024::cogsci_paper
header-includes: \cogscifinalcopy
---

@@

<!--(Note: we should talk about whether it makes more sense to do all the methods and then all the results versus the current expt 1 than expt 2 layout.) -->

Our goal for experiment 1 was to test young children's ability to coordinate on descriptions to abstract shapes that their partner could understand. Young children can be very sensitive to task demands and cognitive load that can hide early abilities (CITATIONS), so we adapted the experimental framework from @leung2024, but further simplified it by reducing the total pool of targets and the number of trials children completed. This experiment was pre-registered at https://osf.io/kcv8j.

@@

Statistical analyses were run in brms [@burkner2018] with weakly informative priors. We present the estimate and 95% credible intervals.

@@

<!--[less important, could drop] --> As another measure of children's performance, we looked at how long each trial took to see if children were getting faster over time. We ran a Bayesian mixed effects model of how long each trial took over time: `r speed_form |> form()`. The first critical trial averaged `r stats_text(speed_mod_1, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_1, 2)`).

@@

In iterated reference games in adults, a canonical finding is that the length of descriptions goes from long to short over repeated references to the initially hard to describe shapes [@clark1986; @hawkins2020b; @boyce2024]. We looked at how long the descriptions the children used were to see if the same trend occurred. 
We ran a Bayesian mixed effects model of how long of a description the "teller" produced: `r description_form |> form()`. The initial length was `r stats_text(description_mod_1, 1)` words and description length was relatively stable over time (change of `r stats_text(description_mod_1, 2)` per trial, shown in Figure \ref{length}). How long of a description is initially warranted depends on how iconic or easily describable the shape is, as well as how large and close the contrast set is. In this case, the low number of options to distinguish and relatively easy to describe shapes may have meant long initial descriptions were less necessary. The adult controls in @leung2024 use shorter initial descriptions than adults in studies with larger arrays of harder to distinguish images. However, young children may also choose to provide shorter descriptions than adults would, and young children may accept shorter descriptions as adequate whereas adults may ask for more information to increase their confidence level before making a guess.

@@

While description length is used as a proxy for measuring convention formation, better measurements for convergence look at the actual content of the utterances, which does not always track description length [@boyce2024]. Following @boyce2024, we use Sentence-BERT to embed the descriptions in a semantic vector space, and then use the cosine between embeddings as a measure of semantic similarity between descriptions [@reimers2019]. We compared the semantic similarities between descriptions of the same target based on who produced the description. Our question was whether the two children within the same game produced descriptions that are more similar to each other than two children in different games.

@@

In the first experiment, we adapted the paradigm of @leung2024 for pairs of children, taking an already simple set up and making it shorter. Our goal was to see if young children were at all able to provide descriptions to each other adequate to identify abstract figures, so children received a lot of scaffolding around the experimental interaction. Sometimes, this included experimenters echoing children's descriptions which could potentially influence children's responses. In Experiment 2, we repeated the same paradigm, with tighter experimenter controls and a larger sample size.

@@

```{=latex}
\begin{table}
	\caption{Example descriptions children used for different target tangram images.}
	\label{example}
	\begin{tabular}[H]{p{16em}p{4em}}
		\hline
		%hold
		$\bullet$ person & 		\multirow{6}{*}{\includegraphics[width=.5in]{I1.jpg}}\\

		$\bullet$ a person holding a sandwich &\\
		$\bullet$ a people carrying a box of dirt &\\
		$\bullet$ a monster &\\
		$\bullet$ someone holding a plate and giving it to a restaurant and has watermelon &\\
		\hline
		$\bullet$ vampire & \multirow{7}{*}{\includegraphics[width=.5in]{E1.jpg}}\\
		$\bullet$ hopping & \\
		$\bullet$ a person flying &\\
		$\bullet$ a person &\\
		$\bullet$ a kite &\\
		$\bullet$ a triangle with a head on it with feet &\\
				\multicolumn{2}{l}{$\bullet$  somebody skydiving, not in the airplane}\\
		\hline
		$\bullet$ racecar & \multirow{7}{*}{\includegraphics[width=.5in]{D1.jpg}}\\
		$\bullet$ airplane &\\
		$\bullet$ alligator &\\
		$\bullet$ a person fell down &\\
		$\bullet$ a boat &\\
		$\bullet$ a person that's in a race car that has one triangle and two triangles &\\
		\hline

		$\bullet$ person & \multirow{4}{*}{\includegraphics[width=.5in]{B1.jpg}}\\
		$\bullet$ a person walking &\\
		$\bullet$ a person looking down &\\
		$\bullet$ a people, but it doesn't have any arms& \\
		\hline
	\end{tabular}
\end{table}

@@

# Additional results

## Combining across experiments

<!-- Might want to split out and incorporate in along with expt 2 analyses? -->

@@

# acc_mod_data_meta |> summarize(m=mean(correct.num))

@@

<!-- Query -- I did not include experiment # as a random effect here, and maybe should have? I can rerun if we decide we want this analysis and think expt # as a random effect with game nested within it would be useful. -->

## Relation between accuracy and convergence

@@

Across 2 experiments and 51 pairs of 4 and 5-year old children, we tested how well children were at producing referential expressions to allow their partner to find a matching abstract shape. Children varied substantially in what sorts of descriptions they produced, but overall accuracy was high (85%), indicating that children were generally able to produce adequate descriptions. Additionally, children's utterances showed signs of converging toward conceptual pacts.  While this task is substantially scaled down relative to measures used for adult competence, and children do not display all the typical trends (i.e no sign of shortening of referential expression), it does suggest that the relevant communication skills are present at least in rudimentary form by the end of the preschool years.
